\documentclass[xcolor=pdftex,dvipsnames,table,mathserif,aspectratio=169]{beamer}
\usetheme{metropolis}
%\usetheme{Darmstadt}
%\usepackage{times}
%\usefonttheme{structurebold}

\usepackage[english]{babel}
%\usepackage[table]{xcolor}
\usepackage{pgf,pgfarrows,pgfnodes,pgfautomata,pgfheaps}
\usepackage{amsmath,amssymb,setspace}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{relsize}
\usepackage[absolute,overlay]{textpos} 
\newenvironment{reference}[2]{% 
  \begin{textblock*}{\textwidth}(#1,#2) 
      \footnotesize\it\bgroup\color{red!50!black}}{\egroup\end{textblock*}} 

\DeclareMathSizes{10}{10}{6}{6} 


\title [Healthcare and IO]{Willingness to Pay and Healthcare}
\author{C.Conlon }
\institute{Grad IO }
\date{Fall 2019}
\setbeamerfont{equation}{size=\tiny}
\begin{document}

\begin{frame}
\titlepage
\end{frame}


\begin{frame}{Healthcare and IO}
In recent years, IO has made inroads into questions related to healthcare:
\begin{itemize}
\item How do hospital (systems) and insurers interact?
\item What is the value of adding a hospital to an insurer's network?
\item What determines the market power of insurers? hospitals?
\item Can steering incentives (in network/out of network) be effective in reducing costs?
\end{itemize}
\end{frame}

\begin{frame}{Today's Reading:}
In recent years, IO has made inroads into questions related to healthcare:
\begin{itemize}
\item Capps, Dranove, Satterthwaite (2003)
\item Ho AER (2012)
\end{itemize}
\end{frame}



\begin{frame}{CDS(2003): Motivation}
The Government keeps losing hospital mergers (basically all of them):
\begin{itemize}
\item Some people travel enormous distances to get treatment at hospitals (Mayo Clinic, Cleveland Clinic, Johns Hopkins, Mass General, etc.)
\item These cases are lost at the stage of \alert{market definition} under the \alert{hypothetical monopolist test}.
\begin{itemize}
\item Recall: SSNIP of 5\% price increase for all products. If this is not profitable -- include next closest substitute. Continue adding products to define relevant market.
\item Resulting markets are so large (and unconcentrated) that mergers always lead to negligible changes in market power
\end{itemize}
\item What if we could measure relevant markets better?
\end{itemize}
\end{frame} 


\begin{frame}{CDS(2003): Ex-Ante and Ex-Post WTP}
What is the value that $i$ places on having hospital $j$ included in network $G$?
\begin{align*}
U_{i j} &=\alpha R_{j}+H_{j}^{\prime} \Gamma X_{i}+\tau_{1} T_{i j}+\tau_{2} T_{i j} \cdot X_{i}+\tau_{3} T_{i j} \cdot R_{j}-\gamma\left(Y_{i}, Z_{i}\right) P_{j}\left(Z_{i}\right)+\varepsilon_{i j} \\ &=U\left(H_{j}, X_{i}, \lambda_{i}\right)-\gamma\left(X_{i}\right) P_{j}\left(Z_{i}\right)+\varepsilon_{i j} 
\end{align*}

\begin{itemize}
\item $H_{j}=\left[R_{j}, S_{j}\right]$ is partitioned into generic $R$ and diagnosis specific $S$ characteristics.
\item $X_{i}=\left[Y_{i}, Z_{i}\right]$ is partitioned into demographic $Y$ and diagnosis specific $Z$ characteristics.
\item $T_{i j}=T_{j}\left(\lambda_{i}\right)$ distance from $i$'s home to hospital $j$
\item This paper: ignore $\gamma\left(X_{i}\right) P_{j}\left(Z_{i}\right)$ since $P_{j'}(Z_i) \approx P_j(z_i)$
\end{itemize}
\end{frame} 



\end{document}













































